---
title: "Getting Started With BUCSS"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started With BUCSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why Not Plan From the Observed Effect Size?

A natural way to plan a new study is to take the effect size from a prior study
and compute the sample size that would give, say, 80% power for an effect of
that size. The trouble is that the prior effect size is rarely a trustworthy
target, for two reasons.

* **Publication bias.** The published literature is filtered. Studies that
  reached statistical significance are far more likely to appear in print than
  those that did not, so the effects you can read about are systematically
  larger than the effects that were actually run. Planning from a published
  estimate inherits that upward bias.
* **Uncertainty.** Even without any filtering, a single study's effect size is
  one draw from a sampling distribution. Treating that one number as if it were
  the population value ignores how much it could have come out differently.

Both forces push the same way: they make the prior effect look bigger and more
certain than it is, so a study planned at face value tends to be underpowered.

The uncertainty piece is easy to underestimate. Even with no publication bias,
the sample size needed for a target power is a nonlinear function of the effect
size, so an overestimate reduces power (by planning too small a study) by more
than an equally sized underestimate raises it (by planning a larger one).
Symmetric error in the estimate therefore produces an asymmetric loss of power,
which is why correcting for uncertainty, not just bias, matters.

## What BUCSS Does

BUCSS ("Bias and Uncertainty Corrected Sample Size") corrects the prior study's
effect before planning. It uses the prior study's test statistic and design to
build the likelihood of a truncated noncentral distribution, truncated because
small, nonsignificant effects go unpublished (Taylor & Muller, 1996), then reads
off a corrected noncentrality parameter at a chosen level of *assurance*. That corrected
noncentrality parameter, rather than the face-value estimate, drives the sample
size calculation for the planned study. The method is described in Anderson,
Kelley, and Maxwell (2017).

```{r}
library(BUCSS)
```

## A First Example

Suppose a prior study used an independent samples *t* test with 20 participants
per group and reported *t* = 3.00. We want to plan a new study with 80% power,
and we want 80% assurance, that is, we want the realized power to reach or
exceed 80% in 80% of replications.

```{r}
result <- ss_buc_independent_t(t_observed = 3, n = 20, alpha_prior = .05,
                         alpha_planned = .05, assurance = .80, power = .80)
result
```

The correction calls for 130 participants per group, far more than a face-value
calculation from *t* = 3.00 would suggest.

## Reading the Result

Every `ss_buc_*` function returns a tidy object that prints the design, the
necessary sample size with the unit it is counted in, the implied total sample
size where the unit is per group or per cell, the bias and uncertainty adjusted
noncentrality parameter, the largest assurance the prior result can support, and
the planning inputs you supplied. Because the object is an ordinary `data.frame`
underneath, you can pull the two quantities out directly:

```{r}
result$value[result$term == "necessary_sample_size"]
result$value[result$term == "ncp_adjusted"]
```

The unit the sample size is counted in depends on the design (per group, per
cell, number of pairs, or total) and travels on the object so you never have to
guess:

```{r}
attr(result, "sample_size_unit")
```

For programmatic use the broom verbs give the convenient shapes, exactly as in
the **DMAR** package: `tidy()` returns a one-row data frame with one column per
quantity (so you pull a value out by name), and `glance()` returns a one-row
summary.

```{r}
tidy(result)
tidy(result)$necessary_sample_size
glance(result)
```

## The Replication Example

When the prior study has unequal group sizes, you can pass both. Here a prior
study reported *t* = 3.00 with group sizes of 50 and 55, and we ask for 90%
assurance:

```{r}
ss_buc_independent_t(t_observed = 3, n = c(50, 55), alpha_prior = .05,
               alpha_planned = .05, power = .80, assurance = .90)
```

The corrected plan calls for 1485 per group, a vivid illustration of how much
publication bias and uncertainty can inflate the sample size needed for a
dependable replication. Anderson et al. (2017) quoted 1482 for this example
under the original grid-search method; as of BUCSS 2.0.0 the adjusted
noncentrality parameter is found by direct root finding, which gives 1485.

## The Function Family

BUCSS covers a range of designs, one function per analysis. Each summarizes the
prior study with a single test statistic plus the design.

| Function | Analysis | Prior statistic |
|----------|----------|-----------------|
| `ss_buc_independent_t` (alias `ss_buc_smd`) | independent *t* test / standardized mean difference | `t_observed` |
| `ss_buc_paired_t` (alias `ss_buc_smd_paired`) | dependent (paired) *t* test | `t_observed` |
| `ss_buc_one_way_anova` | one-way between-subjects ANOVA | `F_observed` |
| `ss_buc_factorial_anova` | two-way between-subjects ANOVA | `F_observed` |
| `ss_buc_factorial_anova_general` | between-subjects ANOVA, any factors or contrast | `F_observed` |
| `ss_buc_rm_anova` | within-subjects ANOVA (one or two-way) | `F_observed` |
| `ss_buc_rm_anova_general` | within-subjects ANOVA, any factors | `F_observed` |
| `ss_buc_mixed_anova` | split-plot (mixed) ANOVA | `F_observed` |
| `ss_buc_mixed_anova_general` | split-plot ANOVA, any factors | `F_observed` |
| `ss_buc_reg_coef` | single regression coefficient | `t_observed` |
| `ss_buc_R2` | omnibus model R-squared | `F_observed` |
| `ss_buc_reg_joint` | joint test of a subset of predictors | `F_observed` |

See `vignette("worked-examples", package = "BUCSS")` for one worked example of
each, `vignette("understanding-assurance", package = "BUCSS")` for how the
`assurance` and `alpha_prior` settings shape the correction,
`vignette("reading-output", package = "BUCSS")` for the print, `tidy()`, and
`glance()` views of the result, and
`vignette("planning-replications", package = "BUCSS")` for a worked
demonstration of why a literature of underpowered studies misleads and how the
correction repairs the replication sample size.

## References

Anderson, S. F., & Kelley, K. (2024). Sample size planning for replication
studies: The devil is in the design. *Psychological Methods, 29*, 844--867.
https://doi.org/10.1037/met0000520

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724

Taylor, D. J., & Muller, K. E. (1996). Bias in linear model power and sample
size calculation due to estimating noncentrality. *Communications in
Statistics: Theory and Methods, 25*, 1595--1610.
