---
title: "Planning a Correlation Study"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Planning a Correlation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

A Pearson correlation is one of the most commonly reported effects in the
social sciences, and one of the most commonly replicated. Suppose a published
study of 100 participants reported a correlation of *r* = .35, and you want to
plan a study powered to find that relationship again. As with every planner in
this package, the reported correlation should not be taken at face value: it
was selected for publication in part because it was large, so planning from it
without correction tends to yield a study that is too small. `ss_buc_correlation()`
applies the bias and uncertainty correction of Anderson, Kelley, and Maxwell
(2017) and returns the total sample size the replication needs.

```{r}
ss_buc_correlation(r_observed = .35, N = 100, alpha_prior = .05,
                   alpha_planned = .05, assurance = .80, desired_power = .80)
```

The returned `necessary_N` is the total sample size; `ncp_adjusted` is the
corrected noncentrality parameter, from which the corrected correlation is
recoverable as `sqrt(ncp / (ncp + N))`. For the meaning of `assurance` and
`alpha_prior`, and how each moves the recommended sample size, see
`vignette("understanding-assurance", package = "BUCSS")`. Briefly, raising
assurance buys protection against an unlucky prior estimate at the cost of a
larger study:

```{r}
assurances <- c(.5, .7, .8, .9)
sizes <- sapply(assurances, function(a) {
  ss_buc_correlation(r_observed = .35, N = 100, assurance = a)$value[1]
})
data.frame(assurance = assurances, necessary_N = sizes)
```

## A Correlation Is Not a Regression Slope

The test of whether a correlation differs from zero is arithmetically the test
of the slope in a simple regression, so it is tempting to plan a correlation
study with the single-coefficient regression planner. Doing so gives a slightly
different, and slightly too small, answer:

```{r}
t_equivalent <- .35 * sqrt(98 / (1 - .35^2))
c(correlation = ss_buc_correlation(t_observed = t_equivalent, N = 100)$value[1],
  regression  = ss_buc_reg_coef(t_observed = t_equivalent, N = 100, p = 1)$value[1])
```

The gap is small here, a few participants, but it is systematic, and the reason
is worth understanding because it decides which planner is correct for your
design.

The regression planner is exact when the predictor is **fixed by design**: the
researcher chooses the values of the predictor, as in an experiment with
assigned doses. A correlation study is different. It **samples both
variables**, and the sampled spread of the predictor is itself random from
study to study. Conditional on the spread that happens to be drawn, the test
statistic is a noncentral *F*, but its noncentrality parameter rises and falls
with that spread, which follows a chi-square distribution on *N* - 1 degrees of
freedom. The distribution that actually governs the statistic is therefore a
mixture of noncentral *F* distributions, not a single one, and a mixture is
more spread out than its center. Planning as though the statistic were a plain
noncentral *F* overstates the planned study's power, by up to about .04, and so
returns a sample size that is a few participants short.

`ss_buc_correlation()` uses the mixture. It has an exact closed form (the
Poisson weights that build the noncentral *F* become negative binomial weights
when mixed over the chi-square), so the returned sample size is exact for
bivariate normal data rather than an approximation. The rule of thumb is
simple:

* If both variables are sampled, as in an observational correlation study, use
  `ss_buc_correlation()`.
* If the predictor is fixed by the design, the effect is a regression slope,
  and `ss_buc_reg_coef()` with `p = 1` is the planner for it.

You may supply the prior result either as the correlation `r_observed` or as
its *t* statistic `t_observed`; the two are equivalent.

## Assuming Less Publication Bias

`alpha_prior` encodes how strongly publication bias is assumed to have shaped
the prior literature. The default of .05 assumes the field publishes results
significant at the .05 level; a larger value assumes a more permissive
literature and therefore a smaller correction. A prior study that escaped the
file drawer entirely, such as a registered report or a pilot, is described by
`alpha_prior = 1`.

```{r}
alphas <- c(.05, .20, 1)
sizes <- sapply(alphas, function(ap) {
  ss_buc_correlation(r_observed = .35, N = 100, alpha_prior = ap)$value[1]
})
data.frame(alpha_prior = alphas, necessary_N = sizes)
```

## Auditing the Plan by Simulation

The plan above rests on a promise about the long run: at `assurance = .80`,
about 80% of studies planned this way should reach the target power. That
promise is easy to state and hard to picture. `ss_buc_sensitivity()` makes it
concrete by simulating the very process the correction models. It draws many
prior studies from a true effect you name, keeps only the ones a literature
would have published, re-plans from each survivor, and reports how often those
plans actually attain the target.

You must name the true effect; there is no default, because letting the method
assume the corrected estimate is true would be asking it to grade its own
work. A natural first choice is the plan's own corrected parameter, which asks
"if the correction happened to be exactly right, how would plans built this
way behave?"

```{r}
plan <- ss_buc_correlation(r_observed = .35, N = 100, assurance = .80)
ncp <- plan$value[plan$term == "ncp_adjusted"]
ss_buc_sensitivity(plan, true_ncp = ncp, replications = 2000, seed = 2017)
```

Two attainment rates are reported, and the difference between them is the point
of the exercise.

* **`attainment_with_refusals`** is the fraction of published prior studies for
  which a plan built this way reaches the target, counting a refusal (a prior
  result too weak to plan from, which the function declines rather than guesses
  at) as a success, because declining is the most cautious plan of all. This is
  the quantity the assurance guarantee is about, and it lands near the
  requested `assurance` of .80.
* **`attainment`** is the same fraction among only the studies that received a
  plan. It is lower, because it excludes the refusals, and it is what a
  researcher who did get a plan actually experiences.

The gap between the two is the refusal rate, reported alongside them. A large
gap is not a flaw in the method; it is the visible price of a method that
refuses to plan from a prior result too weak to support the request, rather
than returning a confident but hollow number. `size_at_true_effect` is the
sample size the true effect really requires, the benchmark each simulated plan
is measured against, and the quartiles show how widely the plans scatter around
it.

The most instructive use of the function is to vary the assumed truth. Ask what
happens if the real correlation is smaller than the corrected estimate, which
is the case a cautious planner most wants to guard against:

```{r}
smaller <- ss_buc_sensitivity(plan, true_ncp = ncp * 0.6, replications = 2000,
                              seed = 2017)
smaller$value[smaller$term %in% c("attainment", "attainment_with_refusals",
                                  "refusal_rate", "size_at_true_effect")]
```

## When the Prior Result Cannot Support the Request

Every prior result has a largest assurance it can support, the closed-form
value `1 - p/alpha_prior`, where *p* is the prior study's *p* value. Ask for
more and the correction drives the noncentrality parameter to zero: the prior
effect cannot be distinguished from nothing at that level of confidence, so no
sample size attains the target and the function stops rather than returning a
meaningless answer. The strong prior used throughout this vignette supports
assurance up to about .99, so this does not arise for it, but a weaker prior
reaches its ceiling quickly. `plot()` on any plan draws the ceiling and the way
the required sample size climbs toward it; see
`vignette("understanding-assurance", package = "BUCSS")` for that figure and a
fuller discussion.

## References

Anderson, S. F., & Kelley, K. (2024). Sample size planning for replication
studies: The devil is in the design. *Psychological Methods, 29*, 844--867.
https://doi.org/10.1037/met0000520

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724

Taylor, D. J., & Muller, K. E. (1996). Bias in linear model power and sample
size calculation due to estimating noncentrality. *Communications in
Statistics: Theory and Methods, 25*, 1595--1610.
