---
title: "Planning Replication Studies"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Planning Replication Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
```

This vignette is adapted from a graduate seminar exercise. It builds up, step by
step, the problem that BUCSS exists to solve: when a research literature is
built from underpowered studies, the effect sizes that get published are
systematically too large, so a replication planned from a published result is
itself underpowered, often badly. We first reproduce that problem with a small
simulation, then show how the bias and uncertainty correction repairs the
sample size plan.

```{r}
library(BUCSS)
```

## A Literature Built From Underpowered Studies

Imagine a true effect that is real but small. Many laboratories run a two-group
experiment to detect it, each with a modest sample size, and (as is the norm)
only the statistically significant results find their way into print. We can
watch this happen by simulating it.

The true standardized mean difference is `d = 0.2` (Cohen's "small"). Each study
uses 30 participants per group. We run 1000 such studies and keep track of which
ones reach `p < .05`.

```{r}
set.seed(2017)

true_d <- 0.2     # the true (small) population effect
n      <- 30      # participants per group in each study
G      <- 1000    # number of studies the field runs

studies <- replicate(G, {
  control   <- rnorm(n, mean = 0,      sd = 1)
  treatment <- rnorm(n, mean = true_d, sd = 1)
  tt <- t.test(treatment, control, var.equal = TRUE)
  # For equal n per group, Cohen's d = t * sqrt(2 / n).
  c(t = unname(tt$statistic), p = tt$p.value,
    d = unname(tt$statistic) * sqrt(2 / n))
})
studies <- as.data.frame(t(studies))
studies$published <- studies$p < .05
```

How often does a single study detect this effect? That is just its statistical
power, and it is low:

```{r}
power.t.test(delta = true_d, n = n, sig.level = .05)$power
mean(studies$published)   # the simulated proportion that "got published"
```

About one study in nine reaches significance. Now compare the average effect
size across *all* studies with the average across only the *published* ones:

```{r}
c(all       = mean(studies$d),
  published = mean(studies$d[studies$published]))
```

The average across all studies sits near the truth of 0.2. The average across
published studies is far larger, because a small study can only clear the
significance filter when its sample effect happens to land high. The filter does
not select for true effects; it selects for large *estimates*. This is
publication bias, and the figure makes it vivid.

```{r, echo = FALSE}
plot(seq_len(G), studies$d,
     pch = ifelse(studies$published, 19, 1),
     col = ifelse(studies$published, "black", "grey70"),
     cex = 0.6,
     xlab = "Study", ylab = "Observed d",
     main = "Publication bias inflates the observed effect")
abline(h = true_d, lty = 3, lwd = 2, col = "red")
abline(h = mean(studies$d), lty = 2, col = "grey40")
abline(h = mean(studies$d[studies$published]), lwd = 2)
legend("topright", bty = "n", cex = 0.9,
       legend = c("published (p < .05)", "not published",
                  "true d = 0.2", "mean of all", "mean of published"),
       pch = c(19, 1, NA, NA, NA),
       lty = c(NA, NA, 3, 2, 1),
       lwd = c(NA, NA, 2, 1, 2),
       col = c("black", "grey70", "red", "grey40", "black"))
```

The published effects (filled points) are a biased sample of the truth: their
average (solid line) is well above the true value (dashed red line). A reader of
this literature sees only the filled points and naturally concludes the effect
is much larger than it is.

## Why Planning at Face Value Underpowers the Replication

Now put yourself in the position of a researcher who wants to replicate one of
these published findings. Suppose the published study you are working from used
25 participants per group and reported `t = 2.52`. Taken at face value, that
implies a healthy effect size:

```{r}
t_obs  <- 2.52
n_prior <- 25
d_face  <- t_obs * sqrt(2 / n_prior)
d_face
```

If you plan a replication for 80% power using that face-value effect, you need
only a modest sample:

```{r}
n_naive <- ceiling(power.t.test(delta = d_face, power = .80, sig.level = .05)$n)
n_naive
```

But the effect was never really that large; it is one of the inflated published
estimates. If the truth is `d = 0.2`, the power your replication actually has at
that sample size is dismal:

```{r}
power.t.test(delta = 0.2, n = n_naive, sig.level = .05)$power
```

You planned for 80% power and bought yourself about 12%. Planning at face value
does not just lose a little power; it leaves the replication almost as
underpowered as the original. A field that plans this way keeps running studies
that cannot succeed.

## The BUCSS Correction

BUCSS breaks this cycle. Instead of trusting the published effect, it treats the
published test statistic as a draw from a *truncated* distribution, truncated
because only significant results were publishable, and corrects for both the
publication bias and the sampling uncertainty before planning. The same prior
study, run through `ss_buc_independent_t()`, asks for a very different sample
size.

We use `assurance = .50` first. An assurance of .50 corrects for publication
bias only: it asks for the sample size that gives the target power on average
across replications.

```{r}
ss_buc_independent_t(t_observed = 2.52, n = 25,
                     alpha_prior = .05, alpha_planned = .05,
                     assurance = .50, power = .80)
```

Even correcting for publication bias alone, BUCSS calls for far more
participants per group than the naive plan of `r n_naive`. The face-value plan
was not cautious; it was wishful.

Raising the assurance above .50 also corrects for *uncertainty*: it asks for the
sample size that will reach the target power not just on average but in a stated
proportion of replications. With a prior study this small, however, asking for
high assurance reveals a hard truth. The original study simply did not carry
enough information, and BUCSS says so rather than returning a falsely precise
number:

```{r, error = TRUE}
ss_buc_independent_t(t_observed = 2.52, n = 25,
                     alpha_prior = .05, alpha_planned = .05,
                     assurance = .80, power = .80)
```

The message is itself the lesson: an underpowered original study cannot anchor a
high-assurance replication. It tells you the largest assurance this prior can
support and the two levers you can pull (lower the assurance, or raise
`alpha_prior` to assume the literature filtered less aggressively), so you can
make an informed, minimal adjustment rather than a blind one. A well-powered
original study would support assurance of .80 or .95 comfortably; this one does
not, and that is worth knowing before you collect data.

## Reading and Using the Result

Every `ss_buc_*` function returns a tidy object. Print it for a readable
summary, or pull the pieces out programmatically:

```{r}
plan <- ss_buc_independent_t(t_observed = 2.52, n = 25, assurance = .50)
tidy(plan)
tidy(plan)$necessary_sample_size
```

The classic dotted name still works for scripts written against earlier
versions of BUCSS, and the first two elements of the result remain the sample
size and the adjusted noncentrality parameter, exactly as they were:

```{r}
legacy <- suppressWarnings(ss.power.it(t.observed = 2.52, N = 50, assurance = .50))
legacy[[1]]   # necessary sample size per group
```

(Calling the dotted name once per session also prints a deprecation note
pointing to `ss_buc_independent_t()`; new code should use the new name.)

## The Bigger Picture

Detecting the effect again is only one possible goal of a replication. Anderson
and Kelley (2024) describe four, each with its own sample size logic: inferring
that the effect exists (the goal above, served by BUCSS), inferring that the
effect is so small it is essentially null (equivalence testing), estimating the
effect's magnitude precisely (accuracy in parameter estimation, as in the MBESS
package), and planning the next study to sharpen a running meta-analytic
estimate. The common thread is that the design choice comes first: decide what a
successful replication would mean, then plan the sample size for that, and never
take a published effect size at face value.

## References

Anderson, S. F., & Kelley, K. (2024). Sample size planning for replication
studies: The devil is in the design. *Psychological Methods, 29*, 844--867.
https://doi.org/10.1037/met0000520

Anderson, S. F., & Maxwell, S. E. (2017). Addressing the "replication crisis":
Using original studies to design replication studies with appropriate
statistical power. *Multivariate Behavioral Research, 52*, 305--324.

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724

Maxwell, S. E., Delaney, H. D., & Kelley, K. (2027). *Designing experiments and
analyzing data: A model comparison perspective* (4th ed.). Routledge.

Taylor, D. J., & Muller, K. E. (1996). Bias in linear model power and sample
size calculation due to estimating noncentrality. *Communications in Statistics:
Theory and Methods, 25*, 1595--1610.
