---
title: "Reading and Using BUCSS Output"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and Using BUCSS Output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

Every `ss_buc_*` function returns the same kind of object: a tidy result of
class `bucss_power`. This vignette shows the three ways to read it, which follow
the same conventions as the **DMAR** package, so the habits carry over.

We will use a two-way between-subjects ANOVA throughout.

```{r}
result <- ss_buc_factorial_anova(F_observed = 5, N = 120, levels_A = 2,
                                 levels_B = 3, effect = "factor_B")
```

## 1. Read It as a Printed Summary

Printing the object shows the aligned `term`/`value` table, exactly as in
**DMAR**: the design results first (the necessary sample size, named for its
unit; the implied total sample size for per-group and per-cell designs; the
achieved power at the returned size; and the bias and uncertainty adjusted
noncentrality parameter), then the rows echoing the planning inputs you
supplied. Footer lines close with the design, the unit the sample size is
counted in, and the largest assurance this prior result can support.

```{r}
result
```

Displayed values are rounded for reading (whole numbers print clean; other
values print at three significant figures); the stored values keep full
precision. Change the display with `options(bucss.digits = )`.

```{r}
options(bucss.digits = 5)
result
options(bucss.digits = 3)  # restore the default
```

## 2. Pull a Single Number Out

The object is an ordinary `data.frame` with a character `term` column and a
numeric `value` column, so you can index it like any data frame. The design
results come first (here `necessary_n_per_cell`, `total_N`, `actual_power`,
and `ncp_adjusted`; the size row is `necessary_n_per_group` or `necessary_N`
in other designs), followed by one row per planning input, so the assumptions
travel with the result through subsetting and `write.csv()`.

```{r}
result$value[result$term == "necessary_n_per_cell"]
result$value[result$term == "total_N"]
```

That works, but the `tidy()` view below is usually tidier for pulling one value
out by name.

## 3. The Broom Verbs: `tidy()` and `glance()`

`tidy()` returns the compact estimate view, the same shape as **DMAR**'s
sample size planners: one row with `term` (`"sample_size"`), `estimate` (the
necessary sample size in the design's unit), and `power` (the achieved power
at that size).

```{r}
tidy(result)
tidy(result)$estimate
```

`glance()` returns the one-row wide view: every quantity and echoed planning
input as a column, plus the design, the effect tested, the unit, the planned
test's degrees of freedom, and the assurance ceiling. It is the convenient
shape for pulling one value out by name or collecting many results into one
table for reporting.

```{r}
glance(result)
```

Both verbs come from the **generics** package and are re-exported by BUCSS, so
they work as soon as the package is loaded, with no extra setup. The full
precision values are always available in the object itself, since only the
printed display rounds. And for a manuscript's method section,
`planning_sentence()` writes the sentence:

```{r}
planning_sentence(result)
```

## What Lives Where

* Everything numeric is a row of the `value` column: the design results (the
  size row named for its unit, `total_N`, `actual_power`, `ncp_adjusted`) and
  the echoed planning inputs. Rows survive any further arithmetic, subsetting,
  or export you do on the object.
* Only non-numeric metadata travels on attributes: the design label, the
  effect tested, the unit, plus the assurance ceiling and the planned test's
  degrees of freedom. All of it is surfaced by the print method and by
  `glance()`.

## References

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724
