---
title: "Reading and using BUCSS output"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and using BUCSS output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

Every `ss_buc_*` function returns the same kind of object: a tidy result of
class `bucss_power`. This vignette shows the three ways to read it, which follow
the same conventions as the **DMAR** package, so the habits carry over.

We will use a two-way between-subjects ANOVA throughout.

```{r}
result <- ss_buc_factorial_anova(F_observed = 5, N = 120, levels_A = 2,
                                 levels_B = 3, effect = "factor_B")
```

## 1. Read it as a printed summary

Printing the object gives a readable summary: the design, the necessary sample
size with the unit it is counted in (and, for per-group and per-cell designs,
the implied total sample size on the same line), the bias and uncertainty
adjusted noncentrality parameter, the largest assurance this prior result can
support, and the planning inputs you supplied.

```{r}
result
```

The displayed noncentrality parameter is rounded for reading; the stored value
keeps full precision. Change the display with `options(bucss.digits = )`.

```{r}
options(bucss.digits = 5)
result
options(bucss.digits = 3)  # restore the default
```

## 2. Pull a single number out

The object is an ordinary `data.frame` with a character `term` column and a
numeric `value` column, so you can index it like any data frame. The rows are
`necessary_sample_size`, `ncp_adjusted`, and, for per-group and per-cell
designs, `total_N`.

```{r}
result$value[result$term == "necessary_sample_size"]
result$value[result$term == "total_N"]
```

That works, but the `tidy()` view below is usually tidier for pulling one value
out by name.

## 3. The broom verbs: `tidy()` and `glance()`

`tidy()` returns a one-row data frame with one column per quantity (the "wide"
view), so a single value comes out by name with no row indexing. It also
includes the planned test's degrees of freedom and the assurance ceiling.

```{r}
tidy(result)
tidy(result)$necessary_sample_size
tidy(result)$total_N
```

`glance()` returns a one-row model-level summary: the design, the effect tested,
the headline sample sizes, the planned test's degrees of freedom, the unit, and
the assurance ceiling. It is the convenient shape for collecting many results
into one table for reporting.

```{r}
glance(result)
```

Both verbs come from the **generics** package and are re-exported by BUCSS, so
they work as soon as the package is loaded, with no extra setup. The full
precision values are always available in the object itself, since only the
printed display rounds.

## What lives where

* The planned-study quantities (`necessary_sample_size`, `total_N`,
  `ncp_adjusted`) are rows of the `value` column, so they survive any further
  arithmetic you do on the object.
* The design label, the effect tested, the unit, the assurance ceiling, the
  planned test's degrees of freedom, and the planning inputs travel on
  attributes and are surfaced by the print method and by `tidy()`/`glance()`.

## References

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724
