---
title: "Assurance and the publication bias correction"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assurance and the publication bias correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

Two settings control how aggressively BUCSS corrects the prior study's effect:
`assurance` handles uncertainty, and `alpha_prior` handles publication bias.
This vignette explains what each one does and shows how the recommended sample
size responds as you turn each knob. The correction is described by Anderson,
Kelley, and Maxwell (2017) and builds on the truncated noncentral distribution
of Taylor and Muller (1996).

## Assurance: correcting for uncertainty

Assurance is the long-run proportion of replications whose realized power meets
or exceeds the target. If you plan for 80% power with `assurance = .80`, then in
about 80% of replications the realized power will be at least 80%.

* `assurance = .5` selects the median of the likelihood. This corrects for
  publication bias only; it does not add any cushion for uncertainty.
* `assurance > .5` selects a lower quantile of the likelihood, which is a
  smaller, more conservative noncentrality parameter, and therefore a larger
  planned sample size. The extra sample size is the price of protecting against
  the chance that the prior estimate was lucky.
* `assurance < .5` is not recommended (the functions warn), because it implies
  under a 50% chance of reaching the desired power.

Watch the recommended per-group sample size climb as assurance rises, for a
prior independent *t* test with *t* = 3.00 and 20 per group:

```{r}
assurances <- c(.5, .6, .7, .8, .9)
n_by_assurance <- sapply(assurances, function(a) {
  r <- ss_buc_indep_t(t_observed = 3, n = 20, assurance = a, power = .80)
  r$value[r$term == "necessary_sample_size"]
})
data.frame(assurance = assurances, necessary_n_per_group = n_by_assurance)
```

The climb is steep. A prior effect this uncertain calls for 130 per group at
80% assurance and over 5000 per group at 90%.

### When the correction cannot be made

Push assurance high enough on a weak or uncertain prior effect and the corrected
noncentrality parameter is driven to zero. That is the method telling you the
prior effect cannot be estimated with confidence at this level of assurance, so
sample size planning is not possible with these settings. The function stops
rather than returning a meaningless answer:

```{r eval = FALSE}
# Not run: this stops with an informative error.
ss_buc_indep_t(t_observed = 3, n = 20, assurance = .95, power = .80)
#> Error: The corrected noncentrality parameter is zero. Please either choose a
#> lower value of assurance and/or a higher value of 'alpha_prior' for the prior
#> study (e.g., accounting for less publication bias).
```

The two remedies are exactly the two knobs: lower the assurance, or assume less
publication bias by raising `alpha_prior`.

## alpha_prior: modeling publication bias

`alpha_prior` is the significance threshold that governs publication in the
prior study's literature. Setting `alpha_prior = .05` assumes the field
publishes results with *p* < .05 and truncates the likelihood at that point. A
larger value assumes a more permissive literature and therefore less severe
bias; `alpha_prior = 1` assumes no publication bias at all, which is appropriate
for a prior study that was never subject to the file drawer, such as a pilot
study.

Note that `alpha_prior` is not a confidence level and is distinct from
`alpha_planned`, which is the Type I error rate of the study you are designing.

Holding assurance at .80, watch the recommended sample size shrink as we assume
less publication bias:

```{r}
alphas <- c(.05, .10, .20, .50, 1)
n_by_alpha <- sapply(alphas, function(ap) {
  r <- ss_buc_indep_t(t_observed = 3, n = 20, alpha_prior = ap, assurance = .80,
                   power = .80)
  r$value[r$term == "necessary_sample_size"]
})
data.frame(alpha_prior = alphas, necessary_n_per_group = n_by_alpha)
```

Assuming the strongest bias (`alpha_prior = .05`) calls for 130 per group;
assuming none (`alpha_prior = 1`) calls for 37.

## Practical guidance

* Make the corrections as minimal as you can defend. Each knob trades a modeling
  assumption for a smaller sample, and overstating either bias or uncertainty
  inflates the study unnecessarily.
* `assurance = .80` is a common, defensible choice. Reserve very high assurance
  for cases where a failed replication is especially costly, and expect the
  required sample size to grow quickly.
* Choose `alpha_prior` to reflect the publishing threshold in the prior study's
  field. Use `alpha_prior = 1` only when you genuinely believe the prior study
  escaped publication bias.

## References

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample size planning for
more accurate statistical power: A method correcting sample effect sizes for
uncertainty and publication bias. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724

Taylor, D. J., & Muller, K. E. (1996). Bias in linear model power and sample
size calculation due to estimating noncentrality. *Communications in
Statistics: Theory and Methods, 25*, 1595--1610.
