---
title: "Assurance and the Publication Bias Correction"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assurance and the Publication Bias Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

Two settings control how aggressively BUCSS corrects the prior study's effect:
`assurance` handles uncertainty, and `alpha_prior` handles publication bias.
This vignette explains what each one does and shows how the recommended sample
size responds as you turn each knob. The correction is described by Anderson,
Kelley, and Maxwell (2017) and builds on the truncated noncentral distribution
of Taylor and Muller (1996).

## Assurance: Correcting for Uncertainty

Assurance is the long-run proportion of replications whose realized power meets
or exceeds the target. If you plan for 80% power with `assurance = .80`, then in
about 80% of replications the realized power will be at least 80%.

* `assurance = .5` selects the median of the likelihood. This corrects for
  publication bias only; it does not add any cushion for uncertainty.
* `assurance > .5` selects a lower quantile of the likelihood, which is a
  smaller, more conservative noncentrality parameter, and therefore a larger
  planned sample size. The extra sample size is the price of protecting against
  the chance that the prior estimate was lucky.
* `assurance < .5` is not recommended (the functions warn), because it implies
  under a 50% chance of reaching the desired power.

Why does uncertainty call for a larger sample at all? Suppose for a moment there
were no publication bias, so the reported effect were equally likely to fall
above or below the true value. The two kinds of error do not cancel, because the
sample size needed for a given level of power is a nonlinear function of the
effect size. An overestimate leads you to plan too small a study and cuts power
sharply, while an equally sized underestimate leads you to plan a larger study
and raises power by less. Averaged over the ways a noisy estimate can come out,
planning at face value loses more power than it gains, so protecting against
uncertainty means shifting the plan toward a larger sample. That is exactly what
assurance above .5 does.

Watch the recommended per-group sample size climb as assurance rises, for a
prior independent *t* test with *t* = 3.00 and 20 per group:

```{r}
assurances <- c(.5, .6, .7, .8, .9)
n_by_assurance <- sapply(assurances, function(a) {
  r <- ss_buc_independent_t(t_observed = 3, n = 20, assurance = a, desired_power = .80)
  r$value[r$term == "necessary_n_per_group"]
})
data.frame(assurance = assurances, necessary_n_per_group = n_by_assurance)
```

The climb is steep. A prior effect this uncertain calls for 130 per group at
80% assurance and over 5000 per group at 90%.

### When the Correction Cannot Be Made

Push assurance high enough on a weak or uncertain prior effect and the corrected
noncentrality parameter is driven to zero. That is the method telling you the
prior effect cannot be estimated with confidence at this level of assurance, so
sample size planning is not possible with these settings. The function stops
rather than returning a meaningless answer, and the error reports the largest
assurance the prior result can support so you know exactly how far to step back:

```{r eval = FALSE}
# Not run: this stops with an informative error.
ss_buc_independent_t(t_observed = 3, n = 20, assurance = .95, desired_power = .80)
#> Error: Sample size planning is not possible with these settings. After
#> correcting the prior study's effect for publication bias and uncertainty
#> at the requested level of assurance, the corrected noncentrality
#> parameter is zero, which means there is not enough information in the
#> prior result to plan the sample size as desired. With this prior result
#> and 'alpha_prior', the largest 'assurance' that still yields a usable
#> plan is about .90; re-run with 'assurance' at or below that value, or
#> raise 'alpha_prior' to lift the ceiling. 'alpha_prior' does not change
#> the prior study, which is what it is; it is your assumption about the
#> publication threshold the study's literature faced, so raising it toward
#> 1 assumes less publication bias to undo (for example, 'alpha_prior = .10'
#> assumes the literature's journals would have published a result
#> significant at the .10 level, a more lenient policy than .05, and
#> 'alpha_prior = 1' assumes no publication bias at all). See Anderson,
#> Kelley, and Maxwell (2017) and Anderson and Kelley (2024).
```

The reported ceiling is exact: the largest assurance a prior result can support
is `1 - p/alpha_prior`, where *p* is the prior study's *p* value. Here that is
about .90, so .95 is out of reach but .90 or below is fine. The two remedies are
exactly the two knobs: lower the assurance to at or below the reported ceiling,
or assume less publication bias by raising `alpha_prior`, which lifts the
ceiling. When the ceiling is already at or below the .5 floor, only raising
`alpha_prior` can help.

Every successful call reports this same ceiling too, printed in the footer as
"Largest supportable assurance", so you can see how much headroom remains
before the correction breaks down. The printed result also gives the implied
total sample size for per-group and per-cell designs, saving you the
multiplication.

## alpha_prior: Modeling Publication Bias

`alpha_prior` is the significance threshold that governs publication in the
prior study's literature. Setting `alpha_prior = .05` assumes the field
publishes results with *p* < .05 and truncates the likelihood at that point. A
larger value assumes a more permissive literature and therefore less severe
bias; `alpha_prior = 1` assumes no publication bias at all, which is appropriate
for a prior study that was never subject to the file drawer, such as a pilot
study.

A dissertation study is a similar case, since it is typically completed and
defended regardless of how the results turn out. Even a pilot study is not
guaranteed to be bias-free, though. If the researchers would have gone on to the
full study only after a promising pilot result, say *p* < .25, then the pilot
faced an informal filter of its own, and an `alpha_prior` between .05 and 1 may
describe it better than either extreme. An intermediate value above .05 also
suits a prior effect that was not the original study's focus, such as a
secondary effect reported alongside a study's main result, which usually faced a
weaker publication filter than the headline effect did.

Note that `alpha_prior` is not a confidence level and is distinct from
`alpha_planned`, which is the Type I error rate of the study you are designing.

Holding assurance at .80, watch the recommended sample size shrink as we assume
less publication bias:

```{r}
alphas <- c(.05, .10, .20, .50, 1)
n_by_alpha <- sapply(alphas, function(ap) {
  r <- ss_buc_independent_t(t_observed = 3, n = 20, alpha_prior = ap, assurance = .80,
                   desired_power = .80)
  r$value[r$term == "necessary_n_per_group"]
})
data.frame(alpha_prior = alphas, necessary_n_per_group = n_by_alpha)
```

Assuming the strongest bias (`alpha_prior = .05`) calls for 130 per group;
assuming none (`alpha_prior = 1`) calls for 37.

## Practical Guidance

* Make the corrections as minimal as you can defend. Each knob trades a modeling
  assumption for a smaller sample, and overstating either bias or uncertainty
  inflates the study unnecessarily.
* `assurance = .80` is a common, defensible choice. Reserve very high assurance
  for cases where a failed replication is especially costly, and expect the
  required sample size to grow quickly.
* Choose `alpha_prior` to reflect the publishing threshold in the prior study's
  field. Use `alpha_prior = 1` only when you genuinely believe the prior study
  escaped publication bias.

## References

Anderson, S. F., & Kelley, K. (2024). Sample size planning for replication
studies: The devil is in the design. *Psychological Methods, 29*, 844--867.
https://doi.org/10.1037/met0000520

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724

Taylor, D. J., & Muller, K. E. (1996). Bias in linear model power and sample
size calculation due to estimating noncentrality. *Communications in
Statistics: Theory and Methods, 25*, 1595--1610.
