---
title: "Worked Examples Across Designs"
author: "Ken Kelley, Samantha F. Anderson, and Scott E. Maxwell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked Examples Across Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BUCSS)
```

This vignette walks through one example of each `ss_buc_*` function. The pattern
is always the same: summarize the prior study with a single test statistic and
its design, and the function returns the necessary planned-study sample size and
the bias and uncertainty adjusted noncentrality parameter (Anderson, Kelley, &
Maxwell, 2017). Every example uses
80% power and 80% assurance unless noted. For the meaning of the `assurance` and
`alpha_prior` settings, see
`vignette("understanding-assurance", package = "BUCSS")`.

## Independent *t* Test

A prior two-group study reported *t* = 3.00 with 20 participants per group. The
returned sample size is per group.

```{r}
ss_buc_independent_t(t_observed = 3, n = 20, alpha_prior = .05, alpha_planned = .05,
               assurance = .80, desired_power = .80)
```

`ss_buc_smd` is the same planner under an effect size name, for when you are
working from a standardized mean difference rather than a raw *t*; prefer
`ss_buc_independent_t` when the prior *t* was computed from unstandardized data.

## Dependent (Paired) *t* Test

A prior within-pairs study of 40 pairs reported *t* = 3.00. The returned sample
size is the number of pairs.

```{r}
ss_buc_paired_t(t_observed = 3, N = 40, alpha_prior = .05, alpha_planned = .05,
                assurance = .80, desired_power = .80)
```

The alias `ss_buc_smd_paired` serves the standardized mean difference framing of
the same comparison.

## Between-Subjects ANOVA

A one-way fully between-subjects design takes the number of levels of the single
factor. Here a four-group design with total *N* = 120 reported *F* = 5.00 for
the omnibus effect. The returned sample size is per group.

```{r}
ss_buc_one_way_anova(F_observed = 5, N = 120, levels_A = 4, alpha_prior = .05,
                     alpha_planned = .05, assurance = .80, desired_power = .80)
```

A two-way fully between-subjects design takes the levels of both factors and the
effect of interest (`factor_A`, `factor_B`, or `interaction`). Here a 2 x 3
design with total *N* = 120 reported *F* = 5.00 for the main effect of Factor B.
The returned sample size is per cell.

```{r}
ss_buc_factorial_anova(F_observed = 5, N = 120, levels_A = 2, levels_B = 3,
                       effect = "factor_B", alpha_prior = .05,
                       alpha_planned = .05, assurance = .80, desired_power = .80)
```

When the effect of interest is not a standard main effect or interaction, or the
design has more than two factors, use the general form and supply the degrees of
freedom directly. The denominator degrees of freedom equal *N* minus the number
of cells when the cell means are the only estimated parameters, so supplying
`df_denominator = N - cells` (here 120 - 6 = 114) reproduces the two-way result
above. A smaller `df_denominator` accounts for covariates or other nuisance
parameters and calls for a larger sample.

```{r}
ss_buc_factorial_anova_general(F_observed = 5, N = 120, cells = 6,
                               df_numerator = 2, df_denominator = 114,
                               alpha_prior = .05, alpha_planned = .05,
                               assurance = .80, desired_power = .80)
```

## Within-Subjects ANOVA

For a fully within-subjects design, supply the factor levels and the effect.
Here a 2 x 3 within-subjects design with *N* = 60 reported *F* = 5.00 for the
main effect of Factor B. The returned sample size is the total number of
subjects.

```{r}
ss_buc_rm_anova(F_observed = 5, N = 60, levels_A = 2, levels_B = 3,
                effect = "factor_B", alpha_prior = .05, alpha_planned = .05,
                assurance = .80, desired_power = .80)
```

The general form takes the numerator degrees of freedom directly and works for
any number of within-subjects factors. This example uses `assurance = .50`,
which corrects for publication bias only.

```{r}
ss_buc_rm_anova_general(F_observed = 6.5, N = 80, df_numerator = 1,
                        alpha_prior = .05, alpha_planned = .05, assurance = .50,
                        desired_power = .80)
```

## Split-Plot (Mixed) ANOVA

A split-plot design crosses a between-subjects factor with a within-subjects
factor. Supply the levels of each and the effect (`between`, `within`, or
`interaction`). Here a design with 2 between-subjects levels and 3
within-subjects levels and *N* = 60 reported *F* = 5.00 for the within-subjects
effect. The returned sample size is per between-subjects cell.

```{r}
ss_buc_mixed_anova(F_observed = 5, N = 60, levels_between = 2, levels_within = 3,
                   effect = "within", alpha_prior = .05, alpha_planned = .05,
                   assurance = .80, desired_power = .80)
```

The general form takes the numerator degrees of freedom and the number of
between-subjects groups directly. The effect is one of `between_only`,
`within_only`, or `between_within`; supply `df_num_within` when the effect
involves a within-subjects component. The returned sample size is per group.

```{r}
ss_buc_mixed_anova_general(F_observed = 5, N = 90, df_numerator = 2,
                           num_groups = 3, effect = "between_only",
                           alpha_prior = .05, alpha_planned = .05,
                           assurance = .80, desired_power = .80)
```

## Multiple Regression

Anderson (2021) is a step-by-step tutorial for planning regression studies with
BUCSS, and is the best starting point for the three regression planners below.

For a single regression coefficient, supply the *t* statistic for that
coefficient, the prior sample size, and the number of predictors `p`. A prior
study with *N* = 150 and 3 predictors reported *t* = 3.00 for one coefficient.
The returned sample size is the total.

```{r}
ss_buc_reg_coef(t_observed = 3, N = 150, p = 3, alpha_prior = .05,
                alpha_planned = .05, assurance = .80, desired_power = .80)
```

For the omnibus test of the model R-squared, supply the model *F* statistic.
Here a prior study with *N* = 150 and 4 predictors reported *F* = 5.00.

```{r}
ss_buc_R2(F_observed = 5, N = 150, p = 4, alpha_prior = .05,
          alpha_planned = .05, assurance = .80, desired_power = .80)
```

For a joint test of a subset of predictors, also supply `p_joint`, the number of
predictors tested together. Here 2 of the 4 predictors are tested jointly.

```{r}
ss_buc_reg_joint(F_observed = 5, N = 150, p = 4, p_joint = 2,
                 alpha_prior = .05, alpha_planned = .05, assurance = .80,
                 desired_power = .80)
```

## Correlation

A prior study reporting a Pearson correlation is planned from the correlation
itself. A study with *N* = 100 reported *r* = .35. The returned sample size is
the total.

```{r}
ss_buc_correlation(r_observed = .35, N = 100, alpha_prior = .05,
                   alpha_planned = .05, assurance = .80, desired_power = .80)
```

The test of a correlation is arithmetically the test of a simple regression
slope, so it is tempting to plan it as one. That is exact only when the
predictor is fixed by design. A correlation study samples both variables, and
then the statistic's noncentrality parameter is itself random, proportional to
the sampled spread of the predictor. The resulting distribution is more
dispersed, and planning as though it were not overstates the power. The
planner uses the exact distribution for sampled variables, so it asks for a
few more participants than the regression planner does:

```{r}
t_equivalent <- .35 * sqrt(98 / (1 - .35^2))
c(correlation = ss_buc_correlation(t_observed = t_equivalent, N = 100)$value[1],
  regression = ss_buc_reg_coef(t_observed = t_equivalent, N = 100, p = 1)$value[1])
```

If your predictor really is fixed by design, the analysis is a regression
slope, and `ss_buc_reg_coef()` with `p = 1` is the planner for it.

## Analysis of Covariance

Covariates cost error degrees of freedom in both the prior study and the
planned one, so the planner takes their number directly. A prior study with
*N* = 120, three groups, and two covariates reported *F* = 6.00 for the group
effect. The returned sample size is per cell.

```{r}
ss_buc_ancova(F_observed = 6, N = 120, cells = 3, n_covariates = 2,
              alpha_prior = .05, alpha_planned = .05, assurance = .80,
              desired_power = .80)
```

With `n_covariates = 0` this reproduces `ss_buc_one_way_anova()` exactly. For a
single-degree-of-freedom contrast among the adjusted means, supply that
contrast's *F* and set `df_numerator = 1`.

## Multivariate Comparison of Two Groups

When two groups are compared on several outcome variables at once, the prior
study's Hotelling *T*-squared (or the *F* it maps to) plans the replication.
A study with *N* = 80 comparing two groups on 3 outcomes reported *F* = 4.50.
The returned sample size is per group.

```{r}
ss_buc_manova(F_observed = 4.5, N = 80, p_variables = 3, alpha_prior = .05,
              alpha_planned = .05, assurance = .80, desired_power = .80)
```

This covers the multivariate hypothesis whose rank is one, which is what a
two-group comparison gives. A hypothesis of higher rank has a set of
noncentrality parameters rather than a single one, and the correction does not
apply to it.

## Nested Model Comparison

An effect tested by the chi-square difference between two nested models, such
as a constrained path in a structural equation model, is planned from that
difference. A prior study with *N* = 250 reported a difference of 9.50 on 1
degree of freedom.

```{r}
ss_buc_chisq_diff(chisq_observed = 9.5, N = 250, df_difference = 1,
                  alpha_prior = .05, alpha_planned = .05, assurance = .80,
                  desired_power = .80)
```

This is for the *difference* test, not for an omnibus model fit chi-square: a
publishable fit statistic is a small one, which inverts the selection the
correction assumes.

## Unequal Variances

When the two groups' variances differ, the Welch test is the appropriate
comparison, and planning it needs the ratio of the standard deviations in
addition to the group sizes. A prior study with 40 and 55 participants, whose
second group was half again as variable as the first, reported *t* = 3.00.

```{r}
ss_buc_welch_t(t_observed = 3, n_1 = 40, n_2 = 55, sd_ratio = 1.5,
               alpha_prior = .05, alpha_planned = .05, assurance = .80,
               desired_power = .80)
```

The assumed ratio matters more than any other input here, so it is worth
seeing how the plan moves across the values you find plausible:

```{r}
ratios <- c(1, 1.5, 2, 2.5, 3)
sapply(ratios, function(r) {
  res <- ss_buc_welch_t(t_observed = 3, n_1 = 40, n_2 = 55, sd_ratio = r)
  res$value[res$term == "necessary_n_per_group"]
})
```

Planning at equal variances and being wrong is expensive: if the true ratio
were 2, that plan would deliver about .43 power rather than .80. Plan for the
least favorable ratio you find credible.

## References

Anderson, S. F. (2021). Using prior information to plan appropriately powered
regression studies: A tutorial using BUCSS. *Psychological Methods, 26*,
513--526. https://doi.org/10.1037/met0000366

Anderson, S. F., & Kelley, K. (2024). Sample size planning for replication
studies: The devil is in the design. *Psychological Methods, 29*, 844--867.
https://doi.org/10.1037/met0000520

Anderson, S. F., Kelley, K., & Maxwell, S. E. (2017). Sample-size planning for
more accurate statistical power: A method adjusting sample effect sizes for
publication bias and uncertainty. *Psychological Science, 28*, 1547--1562.
https://doi.org/10.1177/0956797617723724
